Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-02-05 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes, however, increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
This changing context alters the natural disturbance regime that shape and maintain tree community structure, composition and functioning [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015], so anticipate the fate of tropical forests requires to understand community response to disturbance.
Disturbance initiate ecological succession comprising a suit of changes in community structure and functioning over time, which rate, endpoint an duration depend on the disturbance intensity [@Chazdon2003;@Willig2018].
Regarding tropical forests, community response to disturbance would be secondary succession, starting with an intact soil and a consistent seed bank, from which we aim would be to highlight predictable trajectories and their underlying ecological processes.

Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017], and modifications in species interactions, mainly through modifications of species density [@Chesson2000]. 
For now, post-disturbance response has been largely studied through forest structural parameters, rapid and convenient to measure, such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into community recovery and resilience [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood.
A leading idea on how the impacts of disturbance on community diversity is the Intermediate Disturbance Hypothesis (IDH) [@Connell1978].
The IDH assumes biodiversity changes in relation to disturbance intensity and frequency, with diversity peaks for intermediate intensity and at intermediate time after disturbance. 
At intermediate level of disturbance, the fluctuation of environmental conditions entails deterministic processes favoring both competitively superior species and fast colonizers, hence preventing competitive exclusion [@Shea2004;@Pulsford2016].

The IDH then assumes both predictable trajectories depending on the disturbance intensity, and, without additional disturbance, the recovery of pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Sheil2003;@Willig2018].
In highly diverse tropical forests, however, the validation of the IDH remains controversial [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hubbel2001;@Hugues2007;@Norden2015;@Norden2017].
Empirical evidence generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which might give limited or misleading information on community disturbance response. 
The IDH pattern of diversity peak under an intermediate disturbance regime can indeed result from various ecological processes [@Sheil2003].
To disentangle these various processes and grasp the different facets of community response to disturbance it is crucial to combine a suitable range of approaches [@Shea2004].
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we elucidateed (i) the community taxonomic and functional recovery and (ii) the ecological processes underlying community post-disturbance trajectories and discussed the validity and the extent of the IDH, and (iii) examined community recovery time.

>>>>

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes, however, increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
This changing context alters the natural disturbance regime that shape and maintain tree community structure, composition and functioning [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015].
Anticipate the fate of tropical forests then requires to understand the determinants of community response to disturbance and the underlying processes.
In tropical forests, disturbance is followed by secondary succession that starts with a maintained forest environment.
Post-disturbance successional trajectories follow environmental variability and comprises a suit of changes in community structure and functioning.
This succession is driven by various processes involving the different species responses to disturbance (as colonisation, facilitation, coexistence, etc) [@Lindenmayer2012;@Garcia_florez2017].
The rate, endpoint an duration of these trajectories depend on the intensity of the disturbance: a current aim of tropical ecology is to highlight the linkage between biodiversity and disturbance and infer predictable trajectories [@Chazdon2003a;@Willig2018].

Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017]. 
For now, post-disturbance trajectories has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Models based on observed trajectories then assessed community time to recovery and highlighted the role of forest structure, forest composition, environmental parameters and disturbance characteristics for the determination of post-disturbance trajectories [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood [@Guitet2018;@Molino2001].
There is a need to further analyse community post-disturbance trajectories to assess community recovery and linkage between biodiversity and disturbance.

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events [@Connell1978], and states that community diversity peaks at intermediate level of disturbance.
Post-disturbance fluctuations of environmental conditions are assumed to prevent competitive exclusion in fostering both competitively superior species and fast colonizers [@Shea2004;@Pulsford2016].
In highly diverse tropical forests, however, the validation of the IDH remains controversial [@Hubbell2001;@Fox2013;@Sheil2013] and observations are often blurry compared to theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017].
The IDH pattern of peaked diversity under intermediate disturbance can result from various processes that should be identified to grasp the conditions for the IDH application and the impact on communities, and to anticipate disturbance outcomes [@Sheil2003;@Shea2004].

Empirical evidence of the IDH generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which is limited as neither community changes in species abundance nor in functional characteristics are accounted for [@Mayfield2010;@Villeger2011]. 
The post-disturbance processes involved, though, can only be disentangled in combining different approaches [@Sheil2003;@Shea2004].
More ecological-meaningful analysis would couple taxonomic richness, evenness, and composition, to highlight the changes in species composition and abundance distribution, and hence the changes in fostered species [@Chaudhary2016]. 
Analysis would as well include a functional approach, that accounts for species biological attributes, would include as the trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a].

In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we elucidateed (i) the community taxonomic and functional recovery and (ii) the ecological processes underlying community post-disturbance trajectories and discussed the validity and the extent of the IDH, and (iii) examined community recovery time.

### 2019-02-05 WholePlotTrajectories.tex
intro

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes, however, increase
the pressure on remaining natural forests \citep{Morales-Hidalgo2015}.
This changing context alters the natural disturbance regime that shape
and maintain tree community structure, composition and functioning
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}, so anticipate
the fate of tropical forests requires to understand community response
to disturbance. Disturbance initiate ecological succession comprising a
suit of changes in community structure and functioning over time, which
rate, endpoint an duration depend on the disturbance intensity
\citep{Chazdon2003, Willig2018}. Regarding tropical forests, community
response to disturbance would be secondary succession, starting with an
intact soil and a consistent seed bank, from which we aim would be to
highlight predictable trajectories and their underlying ecological
processes.

Disturbance change both community abiotic and biotic environments,
through modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and modifications in species interactions,
mainly through modifications of species density \citep{Chesson2000}. For
now, post-disturbance response has been largely studied through forest
structural parameters, rapid and convenient to measure, such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into
community recovery and resilience \citep{Herault2018}. Regarding
community diversity and composition, however, post-disturbance
trajectories have not been as thoroughly understood. A leading idea on
how the impacts of disturbance on community diversity is the
Intermediate Disturbance Hypothesis (IDH) \citep{Connell1978}. The IDH
assumes biodiversity changes in relation to disturbance intensity and
frequency, with diversity peaks for intermediate intensity and at
intermediate time after disturbance. At intermediate level of
disturbance, the fluctuation of environmental conditions entails
deterministic processes favoring both competitively superior species and
fast colonizers, hence preventing competitive exclusion
\citep{Shea2004, Pulsford2016}.

The IDH then assumes both predictable trajectories depending on the
disturbance intensity, and, without additional disturbance, the recovery
of pre-disturbance state defined by the dominance of late-successional
species \citep{Rees2001, Sheil2003, Willig2018}. In highly diverse
tropical forests, however, the validation of the IDH remains
controversial \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hubbel2001, Hugues2007, Norden2015, Norden2017}. Empirical
evidence generally rely on the analysis of a single diversity metric,
often the taxonomic richness \citep{Molino2001}, which might give
limited or misleading information on community disturbance response. The
IDH pattern of diversity peak under an intermediate disturbance regime
can indeed result from various ecological processes \citep{Sheil2003}.
To disentangle these various processes and grasp the different facets of
community response to disturbance it is crucial to combine a suitable
range of approaches \citep{Shea2004}. More ecological-meaningful
analysis would first couple taxonomic richness and evenness, which
reveals the changes in species abundance distribution and hence in the
underlying ecological processes \citep{Chaudhary2016}. Analysis would as
well include taxonomic composition, which is crucial for conservation
issues \citep{Lavorel2002, Bellwood2006}. Eventually, including a
functional approach accounting for species biological attributes would
be highly revealing of ecological process involved in community response
to disturbance. The functional trait-based approach indeed directly
links community diversity, composition and redundancy to ecosystem
functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits related to species ecology and
mediating species performance in a given environment \citep{Diaz2005}.
Specifically in tropical forests, the functional approach revealed the
post-disturbance deterministic processes entailing a shift from a
dominance of ``conservative'' slow-growing species dealing with scarce
resources, to a dominance of ``acquisitive'' fast-growing species with
rapid and efficient use of abundant resources
\citep{Reich2014, Herault2011}. This shift is translated into the
trajectories of average community value of key functional traits related
to resource acquisition (leaf and stem traits) and life-history strategy
(seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we elucidateed (i) the community taxonomic and functional
recovery and (ii) the ecological processes underlying community
post-disturbance trajectories and discussed the validity and the extent
of the IDH, and (iii) examined community recovery time.

>>>>

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes, however, increase
the pressure on remaining natural forests \citep{Morales-Hidalgo2015}.
This changing context alters the natural disturbance regime that shape
and maintain tree community structure, composition and functioning
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. Anticipate the
fate of tropical forests then requires to understand the determinants of
community response to disturbance and the underlying processes. In
tropical forests, disturbance is followed by secondary succession that
starts with a maintained forest environment. Post-disturbance
successional trajectories follow environmental variability and comprises
a suit of changes in community structure and functioning. This
succession is driven by various processes involving the different
species responses to disturbance (as colonisation, facilitation,
coexistence, etc) \citep{Lindenmayer2012, Garcia_florez2017}. The rate,
endpoint an duration of these trajectories depend on the intensity of
the disturbance: a current aim of tropical ecology is to highlight the
linkage between biodiversity and disturbance and infer predictable
trajectories \citep{Chazdon2003a, Willig2018}.

Disturbance change both community abiotic and biotic environments,
through modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}. For now, post-disturbance trajectories has
been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and highlighted
the role of forest structure, forest composition, environmental
parameters and disturbance characteristics for the determination of
post-disturbance trajectories \citep{Herault2018}. Regarding community
diversity and composition, however, post-disturbance trajectories have
not been as thoroughly understood \citep{Guitet2018, Molino2001}. There
is a need to further analyse community post-disturbance trajectories to
assess community recovery and linkage between biodiversity and
disturbance.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}, and states that
community diversity peaks at intermediate level of disturbance.
Post-disturbance fluctuations of environmental conditions are assumed to
prevent competitive exclusion in fostering both competitively superior
species and fast colonizers \citep{Shea2004, Pulsford2016}. In highly
diverse tropical forests, however, the validation of the IDH remains
controversial \citep{Hubbell2001, Fox2013, Sheil2013} and observations
are often blurry compared to theoretical expectations
\citep{Hugues2007, Sheil2003, Norden2017}. The IDH pattern of peaked
diversity under intermediate disturbance can result from various
processes that should be identified to grasp the conditions for the IDH
application and the impact on communities, and to anticipate disturbance
outcomes \citep{Sheil2003, Shea2004}.

Empirical evidence of the IDH generally rely on the analysis of a single
diversity metric, often the taxonomic richness \citep{Molino2001}, which
is limited as neither community changes in species abundance nor in
functional characteristics are accounted for
\citep{Mayfield2010, Villeger2011}. The post-disturbance processes
involved, though, can only be disentangled in combining different
approaches \citep{Sheil2003, Shea2004}. More ecological-meaningful
analysis would couple taxonomic richness, evenness, and composition, to
highlight the changes in species composition and abundance distribution,
and hence the changes in fostered species \citep{Chaudhary2016}.
Analysis would as well include a functional approach, that accounts for
species biological attributes, would include as the trait-based approach
indeed directly links community diversity, composition and redundancy to
ecosystem functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}.

In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we elucidateed (i) the community taxonomic and functional
recovery and (ii) the ecological processes underlying community
post-disturbance trajectories and discussed the validity and the extent
of the IDH, and (iii) examined community recovery time.
